Exploring the Utility of Machine Learning Across Varied Data Formats

### 2024-05-24 homework2/paper/paper.tex
finish 3rd section

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}p{3cm}||>{\centering\arraybackslash}p{2cm}|>{\centering\arraybackslash}p{6cm}| }
        \hline
        \multicolumn{3}{|c|}{List of all attributes in the Salary Prediction dataset} \\
        \hline
        Attribute name & Type & Details \\
        \hline
        fnl & numeric & Socio-economic characteristic of the population from which the individual comes \\
        \hline
        hpw & numeric & Number of work hours per week \\
        \hline
        relation & categorical & The type of relationship in which the individual is involved \\
        \hline
        gain & numeric & Capital gain \\
        \hline
        country & categorical & Country of origin \\
        \hline
        job & categorical & The individual's job \\
        \hline
        edu\_int & numeric & Number of years of study \\
        \hline
        years & numeric & Age of the individual \\
        \hline
        loss & numeric & Loss of capital \\
        \hline
        work\_type & categorical & The job's type \\
        \hline
        partner & categorical & The type of partner the individual has \\
        \hline
        edu & categorical & The individual's type of education \\
        \hline
        gender & categorical & Individual's gender \\
        \hline
        race & categorical & Individual's race \\
        \hline
        prod & numeric & Capital production \\
        \hline
        gtype & categorical & Type of employment contract \\
        \hline
        \caption{Salary Prediction Attributes}
        \label{tab:salary_attributes_description}
   \end{longtable}
\end{center}

@@

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{1cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{5}{|c|}{List of all Continuous Numeric Attributes in the Stroke Prediction dataset} \\
        \hline
        & mean\_blood\_ sugar\_level & cardiovascular\_ issues & body\_mass\_ indicator & high\_ blood\_pressure \\
        \hline\hline
        count & 5110.000000 & 5110.000000 & 4909.000000 & 5110.000000 \\
        \hline
        mean & 106.147677 & 0.054012 & 28.893237 & 0.097456 \\
        \hline
        std & 45.283560 & 0.226063 & 7.854067 & 0.296607 \\
        \hline
        min & 55.120000 & 0.000000 & 10.300000 & 0.000000 \\
        \hline
        25\% & 77.245000 & 0.000000 & 23.500000 & 0.000000 \\
        \hline
        50\% & 91.885000 & 0.000000 & 28.100000 & 0.000000 \\
        \hline
        75\% & 114.090000 & 0.000000 & 33.100000 & 0.000000 \\
        \hline
        max & 271.740000 & 1.000000 & 97.600000 & 1.000000 \\
        \hline
    \end{longtable}
    \begin{longtable}{ |>{\centering\arraybackslash}m{1cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{5}{|c|}{List of all Continuous Numeric Attributes in the Stroke Prediction dataset} \\
        \hline
        & years\_old & chaotic\_sleep & analysis\_results & biological\_age\_index \\
        \hline\hline
        count & 5110.000000 & 5110.000000 & 4599.000000 & 5110.000000 \\
        \hline
        mean & 46.568665 & 0.054012 & 323.523446 & 134.784256 \\
        \hline
        std & 26.593912 & 0.226063 & 101.577442 & 50.399352 \\
        \hline
        min & 0.080000 & 0.000000 & 104.829714 & -15.109456 \\
        \hline
        25\% & 26.000000 & 0.000000 & 254.646209 & 96.710581 \\
        \hline
        50\% & 47.000000 & 0.000000 & 301.031628 & 136.374631 \\
        \hline
        75\% & 63.750000 & 0.000000 & 362.822769 & 172.507322 \\
        \hline
        max & 134.000000 & 1.000000 & 756.807975 & 266.986321 \\
        \hline
        \caption{Continuous Numeric Attributes in Stroke Prediction Dataset}
        \label{tab:continuous_numeric_attributes_stroke} \\
   \end{longtable}
\end{center}

@@

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{5cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{3}{|c|}{List of all Discrete Nominal Attributes in the Salary Prediction dataset} \\
        \hline
        & Non-missing count & Unique values count \\
        \hline\hline
        relation & 9999 & 6 \\
        \hline
        country & 9999 & 41 \\
        \hline
        job & 9999 & 14 \\
        \hline
        work\_type & 9999 & 9 \\
        \hline
        partner & 9999 & 7 \\
        \hline
        edu & 9999 & 16 \\
        \hline
        gender & 9199 & 2 \\
        \hline
        race & 9999 & 5 \\
        \hline
        gtype & 9999 & 2 \\
        \hline
        \caption{Discrete Nominal Attributes in Salary Prediction Dataset}
        \label{tab:discrete_nominal_attributes_salary} \\
    \end{longtable}
\end{center}

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{5cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{3}{|c|}{List of all Discrete Nominal Attributes in the Stroke Prediction dataset} \\
        \hline
        & Non-missing count & Unique values count \\
        \hline\hline
        job\_category & 5110 & 5 \\
        \hline
        sex & 5110 & 2 \\
        \hline
        tobacco\_usage & 5110 & 4 \\
        \hline
        married & 4599 &2 \\
        \hline
        living\_area & 5110 & 2 \\
        \hline

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/correlation_matrix_SalaryPrediction_full.png}
    \caption{Correlation matrix for the Salary Prediction dataset}
    \label{fig:correlation_matrix_example_salary}
\end{figure}
\pagebreak
The correlation matrix and Cramér's V matrix provide valuable insights into the
relationships between attributes in the datasets.By examining these matrices, we
can see that Figure \ref{fig:correlation_matrix_example_salary} the \textit{prod}
attribute is highly correlated with the \textit{gain} attribute, while the
\textit{years} attribute is negatively correlated with the \textit{fnl} attribute.

In Figure \ref{fig:correlation_matrix_example_stroke} we can see that the
\textit{cardiovascular\_issues} attribute is highly correlated with the
\textit{chaotic\_sleep} attribute, while the \textit{body\_mass\_indicator} 
attribute is negatively correlated with the \textit{cerebrovascular\_accident}
attribute.

@@

The Cramér's V matrix in Figures \ref{fig:cramer_v_matrix_example_salary} and
\ref{fig:cramer_v_matrix_example_stroke} provides insights into the association
between discrete nominal attributes. For example, in the Salary Prediction dataset,
the \textit{gtype} attribute is strongly associated with the \textit{gender} attribute,
while in the Stroke Prediction dataset, there are no strong associations between
the discrete nominal attributes.

@@

% ------------------------------------------------------------------------------
\section{Data Preprocessing}

>>>>

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}p{3cm}||>{\centering\arraybackslash}p{2cm}|>{\centering\arraybackslash}p{6cm}| }
        \hline
        \multicolumn{3}{|c|}{List of all attributes in the Salary Prediction dataset} \\
        \hline
        Attribute name & Type & Details \\
        \hline
        fnl & numeric & Socio-economic characteristic of the population from which the individual comes \\
        \hline
        hpw & numeric & Number of work hours per week \\
        \hline
        relation & categorical & The type of relationship in which the individual is involved \\
        \hline
        gain & numeric & Capital gain \\
        \hline
        country & categorical & Country of origin \\
        \hline
        job & categorical & The individual's job \\
        \hline
        edu\_int & numeric & Number of years of study \\
        \hline
        years & numeric & Age of the individual \\
        \hline
        loss & numeric & Loss of capital \\
        \hline
        work\_type & categorical & The job's type \\
        \hline
        partner & categorical & The type of partner the individual has \\
        \hline
        edu & categorical & The individual's type of education \\
        \hline
        gender & categorical & Individual's gender \\
        \hline
        race & categorical & Individual's race \\
        \hline
        prod & numeric & Capital production \\
        \hline
        gtype & categorical & Type of employment contract \\
        \hline
        money & categorical & Whether the individual earns more than \$50,000 per year \\
        \hline
        \caption{Salary Prediction Attributes}
        \label{tab:salary_attributes_description}
   \end{longtable}
\end{center}

@@

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{0.95cm}||>{\centering\arraybackslash}m{2.45cm}|>{\centering\arraybackslash}m{2.45cm}|>{\centering\arraybackslash}m{2.45cm}|>{\centering\arraybackslash}m{2.45cm}|>{\centering\arraybackslash}m{2.45cm}| }
        \hline
        \multicolumn{6}{|c|}{List of all Continuous Numeric Attributes in the Stroke Prediction dataset} \\
        \hline
        & mean\_blood\_ sugar\_level & body\_mass\_ indicator & years\_old & analysis\_results & biological\_age\_ index \\
        \hline\hline
        count & 5110.000000 & 4909.000000 & 5110.000000 & 4599.000000 & 5110.000000 \\
        \hline
        mean & 106.147677 & 28.893237 & 46.568665 & 323.523446 & 134.784256 \\
        \hline
        std & 45.283560 & 7.854067 & 26.593912 & 101.577442 & 50.399352 \\
        \hline
        min & 55.120000 & 10.300000 & 0.080000 & 104.829714 & -15.109456 \\
        \hline
        25\% & 77.245000 & 23.500000 & 26.000000 & 254.646209 & 96.710581 \\
        \hline
        50\% & 91.885000 & 28.100000 & 47.000000 & 301.031628 & 136.374631 \\
        \hline
        75\% & 114.090000 & 33.100000 & 63.750000 & 362.822769 & 172.507322 \\
        \hline
        max & 271.740000 & 97.600000 & 134.000000 & 756.807975 & 266.986321 \\
        \hline
        \caption{Continuous Numeric Attributes in Stroke Prediction Dataset}
        \label{tab:continuous_numeric_attributes_stroke} \\
    \end{longtable}
\end{center}

@@

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{5cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{3}{|c|}{List of all Discrete Nominal Attributes in the Salary Prediction dataset} \\
        \hline
        & Non-missing count & Unique values count \\
        \hline\hline
        relation & 9999 & 6 \\
        \hline
        country & 9999 & 41 \\
        \hline
        job & 9999 & 14 \\
        \hline
        work\_type & 9999 & 9 \\
        \hline
        partner & 9999 & 7 \\
        \hline
        edu & 9999 & 16 \\
        \hline
        gender & 9199 & 2 \\
        \hline
        race & 9999 & 5 \\
        \hline
        gtype & 9999 & 2 \\
        \hline
        money & 9999 & 2 \\
        \hline
        \caption{Discrete Nominal Attributes in Salary Prediction Dataset}
        \label{tab:discrete_nominal_attributes_salary} \\
    \end{longtable}
\end{center}

\begin{center}
    \begin{longtable}{ |>{\centering\arraybackslash}m{5cm}||>{\centering\arraybackslash}m{3cm}|>{\centering\arraybackslash}m{3cm}|}
        \hline
        \multicolumn{3}{|c|}{List of all Discrete Nominal Attributes in the Stroke Prediction dataset} \\
        \hline
        & Non-missing count & Unique values count \\
        \hline\hline
        cardiovascular\_issues & 5110 & 2 \\
        \hline
        job\_category & 5110 & 5 \\
        \hline
        sex & 5110 & 2 \\
        \hline
        tobacco\_usage & 5110 & 4 \\
        \hline
        high\_blood\_pressure & 5110 & 2 \\
        \hline
        married & 4599 &2 \\
        \hline
        living\_area & 5110 & 2 \\
        \hline
        chaotic\_sleep & 5110 & 2 \\
        \hline
        cerebrovascular\_accident & 5110 & 2 \\
        \hline

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../plots/correlation_matrix_SalaryPrediction_full.png}
    \caption{Correlation matrix for the Salary Prediction dataset}
    \label{fig:correlation_matrix_example_salary}
\end{figure}
\pagebreak
The correlation matrix and Cramér's V matrix provide valuable insights into the
relationships between attributes in the datasets. By examining these matrices, we
can see that Figure \ref{fig:correlation_matrix_example_salary} the \textit{prod}
attribute is highly correlated with the \textit{gain} attribute, while the
\textit{years} attribute is negatively correlated with the \textit{fnl} attribute.

In Figure \ref{fig:correlation_matrix_example_stroke} we can see that the
\textit{mean\_blood\_sugar\_level} attribute is highly correlated with the
\textit{analysis\_results} attribute, while the \textit{body\_mass\_indicator} 
attribute is negatively correlated with the \textit{analysis\_results} attribute.

@@

The Cramér's V matrix in Figures \ref{fig:cramer_v_matrix_example_salary} and
\ref{fig:cramer_v_matrix_example_stroke} provides insights into the association
between discrete nominal attributes. For example, in the Salary Prediction dataset,
the \textit{gtype} attribute is strongly associated with the \textit{gender} attribute,
while in the Stroke Prediction dataset, the \textit{cardiovascular\_issues} attribute is strongly
associated with the \textit{chaotic\_sleep} attribute.

@@

% ------------------------------------------------------------------------------
\section{Data Preprocessing}
As highlighted in the previous section, high-quality data is the cornerstone of 
effective machine learning models. However, real-world datasets often exhibit 
various imperfections that can impede model performance.  Our exploration of the 
datasets revealed the presence of several such issues, including:

\begin{itemize}
    \item Missing values for specific attributes.
    \item Extreme values (outliers) within certain attributes.
    \item Redundant attributes with high correlation.
    \item Inconsistent value ranges for numeric attributes.
\end{itemize}

These imperfections necessitate data preprocessing, a crucial step aimed at 
transforming the raw data into a clean and consistent format. This section 
delves into the specific data preprocessing techniques employed in this study. 
By addressing these issues, we aim to optimize the data for subsequent machine 
learning algorithms, ultimately enhancing their effectiveness in extracting 
valuable insights.

As a note, all the scripts for data preprocessing are located in the \textit{'preprocessing'}
folder at the root of the project directory.

\subsection{Handling Missing Values}
Missing data, a common issue in real-world datasets, necessitates the application
of imputation procedures to address these missing values.  Imputation techniques 
can be categorized as either univariate or multivariate:

\begin{itemize}
    \item \textit{Univariate Imputation:} This approach focuses solely on the attribute with missing 
    values. Common univariate techniques include replacing missing values with the 
    mean, median, or most frequent value within the attribute. These methods are 
    simple to implement but may not effectively capture the underlying relationships 
    between attributes.
    \item \textit{Multivariate Imputation:} This more sophisticated approach leverages the values 
    of other attributes within a sample to estimate the missing value. Techniques 
    like regression analysis are often employed to establish relationships between 
    the missing attribute and the remaining attributes. Based on these relationships, 
    a predicted value can be imputed for the missing data point.  Multivariate 
    imputation offers a more nuanced approach but requires careful consideration 
    of the relationships between attributes and potential biases in the imputation 
    process.
\end{itemize}

In the \textit{'impute\_values.py'} script, in the \textit{'missing\_values'} function,
we used the \textit{IterativeImputer}
class from the \textit{sklearn.impute} module to apply multivariate imputation to
address missing values in the datasets for continuous numeric attributes. The script
uses the most frequent value strategy for categorical attributes. The imputed datasets
are saved in the same folder as the original datasets, with the prefix
\textit{'preprocessed\_missing\_'}.

\subsection{Outlier Detection and Treatment}
Outliers, data points that deviate significantly from the rest of the dataset,
can adversely affect the performance of machine learning models. Outliers can
skew statistical measures, distort relationships between attributes, and lead to
poor generalization of the model. Detecting and treating outliers is essential
for ensuring the robustness and reliability of the model.

We purpose to impute the outliers using the \textit{IsolationForest} algorithm
from the \textit{sklearn.ensemble} module. The script \textit{'outlier\_detection.py'}
detects outliers in the continuous numeric attributes of the datasets and replaces
them with the imputed values. The preprocessed datasets with imputed outliers are
saved in the same folder as the original datasets, with the prefix
\textit{'preprocessed\_outliers\_'}.

\subsection{Analysis of Attribute Correlations}
As previously discussed, attribute correlations can provide valuable insights
into the relationships between different attributes in the dataset. By identifying
highly correlated attributes, we can eliminate redundant information and reduce
the dimensionality of the data, leading to more efficient model training and
improved interpretability.

We choose to remove highly correlated attributes found in the section of
Exploratory Data Analysis. These attributes are:

\begin{itemize}
    \item \textit{prod}: it is correlated with \textit{gain} in the Salary Prediction dataset.
    \item \textit{analysis\_results}: it is correlated with \textit{body\_mass\_indicator} in the Stroke Prediction dataset.
    \item \textit{gtype}: it is correlated with \textit{gender} in the Salary Prediction dataset.
    \item 
\end{itemize}

The script \textit{'remove\_correlated\_attributes.py'}
removes those attributes from the train dataset and saves the preprocessed dataset
in the same folder as the original datasets, with the prefix \textit{'preprocessed\_correlated\_'}.

\subsection{Normalization and Standardization}
The numerical attributes in the dataset can vary significantly in their value scales. For example, 
some attributes may have values in the thousands, while others have values in the single digits. 
This disparity in scales can significantly affect algorithms like Logistic Regression.

In algorithms like Logistic Regression, which rely on a linear combination of attribute values,
attributes with larger numerical values can disproportionately influence the model. This dominance 
can lead to biased results and reduce the model's effectiveness.

To mitigate this issue, it is essential to standardize the values of the numeric attributes. 
Standardization adjusts the scales of the attributes, ensuring that each one contributes equally 
to the model's predictions. This process improves the performance and accuracy of the model by 
creating a more balanced and fair representation of the data.
